*Supplementary Material 1*

### 2022-03-25 ms/manuscript.Rmd
Update MS
Add TS fossil date plots

# Targets from MS workflow
tar_load(
  c(
    accepted_species,
    bb_coverage_by_rank,
    bb_total_coverage,
    coverage_by_rank,
    crown_ages,
    family_stem_ages,
    fern_monophy_summ_tbl,
    fern_nonmono_gen,
    fossil_ferns_raw,
    gb_species_by_year,
    other_dates,
    plastome_iqtree_log,
    ref_files,
    sanger_iqtree_log,
    sanger_sampling,
    total_coverage,
    treepl_cv_results
  ),
  store = here::here("_targets")
)
```

@@

We recovered older crown ages for ferns and large clades therein, as well as older stem ages for families relative to previous studies (Figures `r figure_num("divtimes")`, `r s_figure_num("du-ages")`).
This is almost certainly due to our use of a completely revised and greatly expanded set of fossil calibration points relative to previous studies, which not only resulted in a more densely constrained tree but also a higher number of fern families with minimum fossil ages.
Several recent studies exploring divergence times across a global fern phylogeny all used a similar set of 24--26 fossil calibration points [@Schuettpelz2007b; @Testo2016a] or secondary calibration points based on studies using this set [@Rothfels2015a].
However, we found inconsistencies in the application of these fossils due to differences in taxonomic concepts between the original fossil publication and the studies in which they were used.
Moreover, our set of calibration points more than doubles the number used by previous studies.
It is likely that this expanded sampling has yielded older estimates for the stem age of many families, whereas previous studies tended to estimate similar ages by re-using a similar, more limited set of calibration points (Figures `r figure_num("divtimes")`, `r s_figure_num("du-ages")`).

Notably, the scenario of fern diversification suggested by our results strongly conflicts with the hypothesis that Polypodiales diversified in the shadow of angiosperms [@Schneider2004d].
Rather, we estimate that polypod diversification occurred coincidentally or even preceding the diversification of angiosperms (`r s_figure("ltt")`).

<<FIXME: add discussion of some specific groups once we have final ages>>

@@

# Make tibble for annotating major clades
clade_tibble <- 
  # Use assess_monophy to get MRCA of each major clade
  # They should all by monophyletic anyways
  select(sanger_sampling, species, major_clade) %>%
  assess_monophy(
    taxon_sampling = .,
    tree = sanger_con_tree_dated,
    tax_levels = "major_clade"
  ) %>%
  get_result_monophy(1) %>%
  # Exclude lycophytes
  filter(!taxon %in% c(
    "Selaginellales", "Lycopodiales", "Isoëtales", "Isoetales")) %>%
  verify(all(monophyly == "Yes")) %>%
  transmute(major_clade = taxon, mrca = parse_number(mrca)) %>%
  left_join(clade_cols_tib, by = "major_clade")
### plot-prep end
```

@@

```{r div-times-make}
# Convert family ages to almost-long format

# Combine dates in almost-long format:
# one row per family/study combination,
# cols for median, high, and low date
all_ages <-
  bind_rows(
    subset_ages(other_dates, "ts"),
    subset_ages(other_dates, "rothfels"),
    select(other_dates, family, median = schuettpelz_best) %>%
      mutate(author = "schuettpelz"),
    transmute(
      family_stem_ages,
      family,
      author = "current",
      median = age
    )
  ) %>%
  left_join(
    unique(select(sanger_sampling, family, major_clade)), by = "family") %>%
  left_join(clade_tibble, by = "major_clade") %>%
  # (some median dates are NA if that family wasn't included in a study)
  assert(not_na, family, major_clade, code, author) %>%
  # Set family as a factor ordered by current age estimate
  left_join(family_stem_ages, by = "family") %>%
  mutate(family = fct_reorder(family, age)) %>%
  select(-age)

# Define shapes and colors for points
ages_shapes <- c(
  current = 16, # circle
  rothfels = 17, # triangle
  schuettpelz = 15, # square
  ts = 18 # diamond
)

# Format names of papers for printing
ages_sampling <-
  tribble(
    ~source, ~year, ~label,
    "schuettpelz", 2009, "Schuettpelz & Pryer 2009",
    "rothfels", 2015, "Rothfels et al. 2015",
    "ts", 2016, "Testo and Sundue 2016",
    "current", 2021, "This study",
  ) %>%
  mutate(source = fct_reorder(source, year))

# Format y-axis text: show major clade in parentheses
family_labs_tbl <-
  all_ages %>%
  select(family, code) %>%
  unique() %>%
  mutate(label = glue("{family} ({code})"))

@@

# Provide names of periods in caption
# Make sure to read and check that they are all there!
period_names <-
  periods %>%
  # Filter to range including Ordovician to Neogene
  filter(min_age < 475) %>%
  filter(min_age > 1) %>%
  arrange(desc(min_age)) %>%
  mutate(label = glue("{abbr} ({name})")) %>%
  pull(label) %>%
  paste(collapse = ", ")

div_times_plot <-
  ggplot(all_ages) +
  geom_pointrange(
    aes(
      x = median, xmin = low, xmax = high,
      y = family,
      color = author, shape = author),
    position = position_dodge(width = 0.5)) +
  scale_shape_manual(
    values = ages_shapes,
    limits = ages_sampling$source, 
    labels = ages_sampling$label) +
  scale_color_manual(
    values = ages_colors,
    limits = ages_sampling$source, 
    labels = ages_sampling$label) +
  scale_y_discrete(labels = family_labs, expand = expansion(add = 0.7)) +
  scale_x_reverse("Age (millions of years)", expand = c(0,0)) +
  coord_geo(
    dat = list("periods", "eras"),
    pos = list("b", "b"),
    abbrv = list(TRUE, FALSE),
    xlim = c(500, 0),
    fill = "transparent",
    color = "grey20",
    lab_color = "grey30",
    height = unit(1, "lines"),
    expand = TRUE,
    size = 3) +
  guides(shape = guide_legend(nrow = 2)) +
  theme_bw() +
  theme(
    axis.title.y = element_blank(),
    legend.title = element_blank(),
    legend.position = "bottom"
  )
```

>>>>

# Targets from MS workflow
tar_load(
  c(
    accepted_species,
    bb_coverage_by_rank,
    bb_total_coverage,
    coverage_by_rank,
    crown_ages,
    family_stem_ages,
    fern_monophy_summ_tbl,
    fern_nonmono_gen,
    fossil_ferns_raw,
    ftol_ts_comp_mod,
    gb_species_by_year,
    other_dates,
    plastome_iqtree_log,
    ref_files,
    sanger_iqtree_log,
    sanger_sampling,
    total_coverage,
    treepl_cv_results
  ),
  store = here::here("_targets")
)
```

@@

```{r calibration-stats}
ftol_ts_comp_stats <-
ftol_ts_comp_mod %>%
  broom::glance()
```

We recovered older crown ages for ferns and large clades therein, as well as older stem ages for families relative to previous studies (Figures `r figure_num("divtimes")`, `r s_figure_num("du-ages")`).
This is almost certainly due to our use of a completely revised and greatly expanded set of fossil calibration points relative to previous studies, which not only resulted in a more densely constrained tree but also a higher number of fern families with minimum fossil ages.
Our set of fossil calibration points did not add any extremely old fossils (> 200 Ma) that would be expected to strongly push back ages across the tree [save for stem Marattiaceae, which is well-known for its extensive fossil record\; @Rothwell2018]; rather, the vast majority of newly added calibration points are younger than 150 Ma (`r s_figure("comp-fossils")`).

Several recent studies exploring divergence times across a global fern phylogeny all used a similar set of 24--26 fossil calibration points [@Schuettpelz2007b; @Testo2016a] or secondary calibration points based on studies using this set [@Rothfels2015a].
However, we found inconsistencies in the application of these fossils due to differences in taxonomic concepts between the original fossil publication and the studies in which they were used.
Moreover, our set of calibration points more than doubles the number used by previous studies.
It is likely that this expanded sampling has yielded older estimates for the stem age of many families, whereas previous studies tended to estimate similar ages by re-using a similar, more limited set of calibration points (Figures `r figure_num("divtimes")`, `r s_figure_num("du-ages")`).
To test this hypothesis, we conducted additional analysis using the fossil constraints of @Testo2016a but otherwise the exact same methods (FTOL analyzed with treePL).
The resulting stem family ages show a much closer agreement with @Testo2016a (`r r2` = `r ftol_ts_comp_stats$r.squared %>% number(accuracy = 0.01)`, `r pval` = `r ftol_ts_comp_stats$p.value %>% sci_number()`, linear model; Figures `r s_figure_num("comp-stem-ages")`, `r s_figure_num("comp-stem-points")`), indicating that our expanded set of fossil calibration points, not differences in topology or dating methodology, is responsible for the older ages observed in the current study.

Notably, the scenario of fern diversification suggested by our results conflicts with the hypothesis that Polypodiales diversified "in the shadow of angiosperms" [@Schneider2004d].
Rather, we estimate that polypod diversification occurred coincidentally or even preceding the diversification of angiosperms (`r s_figure("ltt")`).
Our estimated age of (`r crown_ages$polypodiales` ma) for crown Polypodiales is considerably older than other recent studies [@Testo2016a; @Du2021] and the fossil record, which only dates back to the Early Cretaceous [@Schneider2001; @Chen1997; @Deng2002; @Schneider2016c; @Regalado2018a].
Molecular ages that are significantly older than the fossil record should be treated with caution; yet, our study is in agreement with others in suggesting a "long fuse" between initial appearance of Polypodiales their subsequent diversification and widespread preservation in the fossil record [@Testo2016a; @Du2021].

Due to the large size of our dataset, carrying out more detailed molecular dating analyses (`r eg`, Bayesian analysis) is computationally difficult with currently available methods [`r eg`, BEAST\; @Drummond2007; @Bouckaert2014].
Here, we have prioritized computational speed and simplicity, since we anticipate re-running the pipeline on a regular basis.
We therefore consider our dated tree as a starting point, and not the final word, for a re-evaluation of divergence times in ferns.
Future studies should focus on utilizing our greatly expanded fern fossil dataset to conduct more thorough molecular dating analyses, possibly including alternative schemes for the age of the root and testing the effects of parameters used for setting priors.

@@

# Make tibble for annotating major clades
clade_tibble <- 
  # Use assess_monophy to get MRCA of each major clade
  # They should all by monophyletic anyways
  select(sanger_sampling, species, major_clade) %>%
  assess_monophy(
    taxon_sampling = .,
    tree = sanger_con_tree_dated,
    tax_levels = "major_clade"
  ) %>%
  get_result_monophy(1) %>%
  # Exclude lycophytes
  filter(!taxon %in% c(
    "Selaginellales", "Lycopodiales", "Isoëtales", "Isoetales")) %>%
  verify(all(monophyly == "Yes")) %>%
  transmute(major_clade = taxon, mrca = parse_number(mrca)) %>%
  left_join(clade_cols_tib, by = "major_clade")

### Prep for stem family age plots ###

# Define shapes and colors for points for stem family age plots
ages_shapes <- c(
  current = 16, # circle
  rothfels = 17, # triangle
  schuettpelz = 15, # square
  ts = 18 # diamond
)

# Format names of papers for printing for stem family age plots
ages_sampling <-
  tribble(
    ~source, ~year, ~label,
    "schuettpelz", 2009, "Schuettpelz & Pryer 2009",
    "rothfels", 2015, "Rothfels et al. 2015",
    "ts", 2016, "Testo and Sundue 2016",
    "current", 2021, "This study",
  ) %>%
  mutate(source = fct_reorder(source, year))

# Provide names of periods in captions
# Make sure to read and check that they are all there!
period_names <-
  periods %>%
  # Filter to range including Ordovician to Neogene
  filter(min_age < 475) %>%
  filter(min_age > 1) %>%
  arrange(desc(min_age)) %>%
  mutate(label = glue("{abbr} ({name})")) %>%
  pull(label) %>%
  paste(collapse = ", ")

### plot-prep end
```

@@

```{r div-times-make}
# Combine family stem ages into almost-long format:
# one row per family/study combination,
# cols for median, high, and low date
all_family_stem_ages <-
  bind_rows(
    subset_ages(other_dates, "ts"),
    subset_ages(other_dates, "rothfels"),
    select(other_dates, family, median = schuettpelz_best) %>%
      mutate(author = "schuettpelz"),
    transmute(
      family_stem_ages,
      family,
      author = "current",
      median = age
    )
  ) %>%
  left_join(
    unique(select(sanger_sampling, family, major_clade)), by = "family") %>%
  left_join(clade_tibble, by = "major_clade") %>%
  # (some median dates are NA if that family wasn't included in a study)
  assert(not_na, family, major_clade, code, author) %>%
  # Set family as a factor ordered by current age estimate
  left_join(family_stem_ages, by = "family") %>%
  mutate(family = fct_reorder(family, age)) %>%
  select(-age)

# Format y-axis text: show major clade in parentheses
family_labs_tbl <-
  all_family_stem_ages %>%
  select(family, code) %>%
  unique() %>%
  mutate(label = glue("{family} ({code})"))

@@

# Make plot
div_times_plot <-
  ggplot(all_family_stem_ages) +
  geom_pointrange(
    aes(
      x = median, xmin = low, xmax = high,
      y = family,
      color = author, shape = author),
    position = position_dodge(width = 0.5)) +
  scale_shape_manual(
    values = ages_shapes,
    limits = ages_sampling$source, 
    labels = ages_sampling$label) +
  scale_color_manual(
    values = ages_colors,
    limits = ages_sampling$source, 
    labels = ages_sampling$label) +
  scale_y_discrete(labels = family_labs, expand = expansion(add = 0.7)) +
  scale_x_reverse("Age (millions of years)", expand = c(0,0)) +
  coord_geo(
    dat = list("periods", "eras"),
    pos = list("b", "b"),
    abbrv = list(TRUE, FALSE),
    xlim = c(500, 0),
    fill = "transparent",
    color = "grey20",
    lab_color = "grey30",
    height = unit(1, "lines"),
    expand = TRUE,
    size = 3) +
  guides(shape = guide_legend(nrow = 2)) +
  theme_bw() +
  theme(
    axis.title.y = element_blank(),
    legend.title = element_blank(),
    legend.position = "bottom"
  )
```

### 2022-03-25 ms/si.Rmd
Update MS
Add TS fossil date plots

```{r, formatters-si}
```

@@

# Targets from MS workflow
tar_load(
  c(
    du_dates_all,
    family_stem_ages,
    fern_monophy_summ_tbl,
    fern_nonmono_gen,
    gb_species_by_year,
    other_dates,
    pteridocat_ppgi_diff_notes,
    pteridocat_ppgi_diff_notes_file,
    ref_files,
    spermato_trees,
    accepted_species
  ),
  store = here::here("_targets")
)
```

@@

@@

<!-- Run the plot prep code -->
```{r plot-prep-main}
```

@@

>>>>

```{r run-formatters-si, eval = params$doc_type == "manual"}
source_chunk("formatters-si")
```

```{r formatters-si, eval = params$doc_type != "manual"}
```

@@

# Targets from MS workflow
tar_load(
  c(
    du_dates_all,
    family_stem_ages,
    fern_monophy_summ_tbl,
    fern_nonmono_gen,
    gb_species_by_year,
    other_dates,
    pteridocat_ppgi_diff_notes,
    pteridocat_ppgi_diff_notes_file,
    ref_files,
    spermato_trees,
    ts_family_stem_ages,
    accepted_species,
    ftol_ts_comp_mod
  ),
  store = here::here("_targets")
)
```

@@

```{r run-plot-prep, eval = params$doc_type == "manual"}
source_chunk("plot-prep-main")
```

```{r plot-prep-main, eval = params$doc_type != "manual"}
```

@@

@@

```{r comp-fossils, fig.width = 7, fig.height = 9, eval = params$doc_type != "manual"}
ts_fossil_calibration_tips %>%
  select(minimum_age, node_calibrated) %>%
  mutate(dataset = "ts") %>%
  bind_rows(
    con_fossil_calibration_tips %>%
      select(minimum_age, node_calibrated) %>%
      mutate(dataset = "current")
  ) %>%
  mutate(node_calibrated = fct_reorder(node_calibrated, minimum_age)) %>%
  ggplot(aes(x = minimum_age, y = node_calibrated, fill = dataset)) +
  geom_col(position = position_dodge()) +
  scale_x_continuous("Age (millions of years)", expand = expansion(mult = c(0,0.1))) +
  theme_bw() +
  scale_fill_manual(
    values = c(current = "grey30", ts = "grey80"),
    labels = c(current = "This Study", ts = "Testo and Sundue 2016")
  ) +
  theme(
    axis.title.y = element_blank(),
    legend.title = element_blank(),
    legend.position = "bottom"
  )
```

**`r s_figure("comp-fossils")`**.
Age of nodes constrained by fossils in this study vs. @Testo2016a. Wide bars indicate the node was only constrained in one study; narrow bars indicate the the node was constrained in both studies.

```{r comp-stem-ages, fig.width = 7, fig.height = 9, eval = params$doc_type != "manual"}
# Combine family stem ages into almost-long format:
# one row per family/study combination,
# cols for median, high, and low date
all_family_stem_ages <-
  bind_rows(
    subset_ages(other_dates, "ts"),
    transmute(
      ts_family_stem_ages,
      family,
      author = "current",
      median = age
    )
  ) %>%
  left_join(
    unique(select(sanger_sampling, family, major_clade)), by = "family") %>%
  left_join(clade_tibble, by = "major_clade") %>%
  # (some median dates are NA if that family wasn't included in a study)
  assert(not_na, family, major_clade, code, author) %>%
  # Set family as a factor ordered by current age estimate
  left_join(ts_family_stem_ages, by = "family") %>%
  mutate(family = fct_reorder(family, age)) %>%
  select(-age)

# Format y-axis text: show major clade in parentheses
family_labs_tbl <-
  all_family_stem_ages %>%
  select(family, code) %>%
  unique() %>%
  mutate(label = glue("{family} ({code})"))

# Make plot
ggplot(all_family_stem_ages) +
  geom_pointrange(
    aes(
      x = median, xmin = low, xmax = high,
      y = family,
      color = author, shape = author),
    position = position_dodge(width = 0.5)) +
  scale_shape_manual(
    values = ages_shapes,
    limits = ages_sampling$source, 
    labels = ages_sampling$label) +
  scale_color_manual(
    values = ages_colors,
    limits = ages_sampling$source, 
    labels = ages_sampling$label) +
  scale_y_discrete(labels = family_labs, expand = expansion(add = 0.7)) +
  scale_x_reverse("Age (millions of years)", expand = c(0,0)) +
  coord_geo(
    dat = list("periods", "eras"),
    pos = list("b", "b"),
    abbrv = list(TRUE, FALSE),
    xlim = c(500, 0),
    fill = "transparent",
    color = "grey20",
    lab_color = "grey30",
    height = unit(1, "lines"),
    expand = TRUE,
    size = 3) +
  guides(shape = guide_legend(nrow = 2)) +
  theme_bw() +
  theme(
    axis.title.y = element_blank(),
    legend.title = element_blank(),
    legend.position = "bottom"
  )
```

**`r s_figure("comp-stem-ages")`**.
Comparison of stem age of fern families (Ma) estimated by @Testo2016a and those estimated using methodology (treePL) and phylogeny (FTOL) of the current study with the fossil constraints of @Testo2016a.
For @Testo2016a, error bars indicate lower and upper 95% highest posterior density levels and point indicates median
For this study, point indicates best (most likely) estimate.
Codes in parentheses after family names indicate major clade as in `r figure("ftol")`.
Period name abbreviations as in `r figure("divtimes")`.

```{r comp-stem-points, fig.width = 6, fig.height = 5, eval = params$doc_type != "manual"}
# For checking that model data is same as plot data
mod_fitted_data <-
ftol_ts_comp_mod %>%
  broom::augment() %>%
  arrange(ftol)

mod_fitted_data_ts <-
ftol_ts_comp_mod %>%
  broom::augment() %>%
  arrange(ts)

# Make dataset for plotting
comp_plot_data <-
all_family_stem_ages %>%
  filter(author == "ts") %>%
  select(family, ts = median, ts_high = high, ts_low = low) %>%
  inner_join(
    filter(all_family_stem_ages, author == "current") %>%
      select(family, ftol = median),
    by = "family"
  ) %>%
  # Double check that model data is same as plot data
  arrange(ftol) %>%
  verify(all(.$ftol == mod_fitted_data$ftol)) %>%
  arrange(ts) %>%
  verify(all(.$ts == mod_fitted_data_ts$ts))

# Extract R2 and P-value for linear mod
stats <- ftol_ts_comp_mod %>%
  broom::glance() %>%
  mutate(label = glue::glue(
    "*R*<sup>2</sup> = {number(r.squared, accuracy = 0.01)}\n
    {pval} = {sci_number(p.value)}"
  )) %>%
  pull(label)

# Extract fitted line for linear mod
fitted <-
broom::tidy(ftol_ts_comp_mod) %>%
  mutate(term = janitor::make_clean_names(term)) %>%
  split(.$term)

ggplot(comp_plot_data, aes(y = ts, x = ftol)) +
  geom_errorbar(aes(ymax = ts_high, ymin = ts_low)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color = "grey60", lty = 2) +
  geom_abline(
    intercept = fitted$intercept$estimate,
    slope = fitted$ts$estimate, color = "grey20"
  ) +
  theme_bw() +
  scale_x_continuous(expand = expansion(mult = c(0, 0.1))) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(x = "This study", y = "Testo and Sundue (2016)") +
  annotate(
    "richtext", x = 0, y = Inf, 
    label.color = NA,
    label = stats, hjust = 0, vjust = 1)
```

**`r s_figure("comp-stem-points")`**.
Comparison of stem age of fern families estimated by @Testo2016a and those estimated using methodology (treePL) and phylogeny (FTOL) of the current study with the fossil constraints of @Testo2016a.
All units in millions of years (Ma).
For @Testo2016a, error bars indicate lower and upper 95% highest posterior density levels and point indicates median.
For this study, point indicates best (most likely) estimate.
Dashed indicates 1:1 relationship.
Solid line indicates linear model fit.
